---
title: "Predictive Analytics - Discussion 2"
author: "Eric Kenney"
date: "2023-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, results=FALSE, message=FALSE}
# Required Libraries
library(fpp3)
library(rstudioapi)
```

## Import Data and Build Initial Plots

```{r imports}
# Read in the data
temps = read.csv(file.path(dirname(documentPath()), "global_climate_data.csv"))

# Filter down to Oregon
temps_or = temps %>% 
  filter(State == "Oregon") %>% 
  select(-c(Country, AverageTemperatureUncertainty, State))

# Convert to tsibble
temps_or = temps_or %>% 
  mutate(Month = yearmonth(dt)) %>% 
  select(-dt) %>% 
  as_tsibble(index = Month)
```

## Reduce Data to Complete Set

There is a noticeable gap prior to 1850 so I filter all data before that period

```{r plot1, echo=FALSE, message=FALSE}
autoplot(temps_or)
```

```{r clean}
# Filter to after 1850 to remove the NAs
temps_or = temps_or %>% 
  filter(year(Month) >= 1850)

# Confirm all NAs are gone
sum(is.na(temps_or$AverageTemperature))
```

## Model and Evaluate

We can see there are no more missing values in the data. No we will build our training and test sets to begin building and evaluating models.

```{r training and test sets}
# Split Training and Test Sets (80/20 Split)
total_obs = dim(temps_or)[1]
train_obs = total_obs * 0.8
test_obs = total_obs - train_obs
train_or = head(temps_or, train_obs)
test_or = tail(temps_or, test_obs)
```

### ETS (A,A,A)

I let the ETS() function automatically pick the model based on AIC and it chose (A,A,A).

```{r AAA, warning=FALSE}
ets_auto = train_or %>% 
  model(ETS(AverageTemperature))

report(ets_auto)

components(ets_auto) %>% 
  autoplot()

# Forecast
ets_auto_forecast = ets_auto %>% 
  forecast(h = test_obs)

accuracy(object = ets_auto_forecast, data = temps_or)

ets_auto_forecast %>% 
  autoplot(test_or)
```

### ETS (A,A,N)

For the next two models I wanted to see if there was significant importance weighted towards either trend or season. For this model I changed season() to "none."

```{r AAN, warning=FALSE}
ets_aan = train_or %>% 
  model(ETS(AverageTemperature ~ error("A") + trend("A") + season("N")))

report(ets_aan)

components(ets_aan) %>% 
  autoplot()

# Forecast
ets_aan_forecast = ets_aan %>% 
  forecast(h = test_obs)

accuracy(object = ets_aan_forecast, data = temps_or)

ets_aan_forecast %>% 
  autoplot(test_or)
```

### ETS (A,N,A)

For the final model I evaluated trend() as "none."

```{r ANA, warning=FALSE}
ets_ana = train_or %>% 
  model(ETS(AverageTemperature ~ error("A") + trend("N") + season("A")))

report(ets_ana)

components(ets_ana) %>% 
  autoplot()

# Forecast
ets_ana_forecast = ets_ana %>% 
  forecast(h = test_obs)

accuracy(object = ets_ana_forecast, data = temps_or)

ets_ana_forecast %>% 
  autoplot(test_or)
```

