---
title: "Predictive Analytics & Forecasting - Homework #3"
author: "Eric Kenney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, results="hide", message=FALSE}
# Required Libraries
require(fpp3)
require(knitr)
require(rstudioapi)

path = dirname(documentPath())
```

## Data Set

For this homework I will be using the Federal Reserve Economic Data (FRED) set on the effective federal interest rate. This includes daily data from July 1954 to August 2023. Additionally, I am using the CPI growth rate information from FRED as a predictor for the interest rate, since the Federal Reserve will raise rates as a way to curb inflation. The inflation data is complete from Mar 1967 on so I filter both sets of data to after that period.

```{r import, warning=FALSE}
# Read in data
interest_rate = read.csv(file.path(path, "federal_interest_rate.csv"))
inflation = read.csv(file.path(path, "inflation.csv"))

# Convert to monthly tsibble
## Interest Rate
interest_rate = interest_rate %>% 
  rename("date" = "DATE", "interest" = "DFF") %>% 
  mutate(date = yearmonth(date)) %>% 
  group_by(date) %>% 
  summarise(interest = mean(interest)) %>% 
  as_tsibble(index = date)

## Inflation
inflation = inflation %>% 
  rename("date" = "DATE", "CPI" = "CPALTT01USM657N") %>% 
  mutate(date = yearmonth(date)) %>% 
  as_tsibble(index = date)

# Plot
## Interest Rate
interest_rate %>% 
  autoplot(interest) +
  labs(x = "Date", y = "Interest Rate (%)",
       title = "Federal Funds Effective Rate",
       subtitle = "Jul 1954 - Aug 2023")

## Inflation
inflation %>% 
  autoplot(CPI) +
  labs(x = "Growth Rate (Previous Period)", y = "Date", 
       title = "CPI Growth Rate in the United States",
       subtitle = "Jan 1960 - Apr 2023")

# Filter down to complete data and join
interest_rate = interest_rate %>% 
  filter(as.Date(date) >= "1967-03-01")

inflation = inflation %>% 
  filter(as.Date(date) >= "1967-03-01")

interest_rate_data = left_join(interest_rate, inflation, by = "date")

# Create Lagged values for the predictor
interest_rate_data = interest_rate_data %>% 
  mutate(CPI_lag = lag(CPI, 12)) %>% 
  filter(!is.na(CPI_lag))
```

The full plot of the data shows a historical high of almost 20% during the recession in the 1980s. The last year (which I intend to forecast) is after a historic drop in rates due to the COVID-19 pandemic. I will reduce the data to end in December 2019 and use 2019 as the test set.

```{r decomp}
# New Time Slice
interest_rate_data = interest_rate_data %>% 
  filter(as.Date(date) <= "2019-12-01")

# STL Decomp of the data
interest_rate_data %>% 
  model(STL(interest)) %>% 
  components() %>% 
  autoplot() +
  labs(x = "Date", y = "Interest Rate (%)",
       title = "STL Decomp of the Federal Funds Effective Rate",
       subtitle = "Mar 1968 - Dec 2019")

# Create Training and Test Sets
total_obs = dim(interest_rate_data)[1]
test_obs = 12
train_obs = total_obs - test_obs

interest_rate_train = head(interest_rate_data, train_obs)
interest_rate_test = tail(interest_rate_data, test_obs)
```

This time slice works better than the most recent five years. The STL decomposition shows a fairly continuous trend with a drop the last year which may cause some issues with the forecasting. I will work with and model from this set.

## Model

I will create three models: ETS, ARIMA, NNETAR, and an ensemble model. 

### ETS

```{r ets, warning=FALSE, message=FALSE}
# Create model
interest_rate_ets = interest_rate_train %>% 
  model(ETS(interest))

# Print report
report(interest_rate_ets)

# Plot components and residuals
interest_rate_ets %>% 
  components() %>% 
  autoplot() + 
  labs(title = "ETS Model (A,A,N) Components")

interest_rate_ets %>% 
  gg_tsresiduals(lag_max = 12) +
  labs(title = "ETS Model (A,A,N) Residuals")

# Ljung-Box test
augment(interest_rate_ets) %>% 
  features(.innov, ljung_box, lag = 12)
```

Fable picked a (A,A,N), a Holt-Winter's Linear Trend. The AIC is 2335 and the residuals look biased to negative and some autocorrelation. A Ljung-Box test gives a p-value of 0.01.

### ARIMA

```{r arima, warning=FALSE, message=FALSE}
# Create model
interest_rate_arima = interest_rate_train %>% 
  model(ARIMA(interest ~ CPI_lag))

# Print report
report(interest_rate_arima)

# Plot residuals
interest_rate_arima %>% 
  gg_tsresiduals(lag_max = 12) +
  labs(title = "ARIMA Model (2,1,0)(0,0,1)[12] Residuals")

# Ljung-Box test
augment(interest_rate_arima) %>% 
  features(.innov, ljung_box, lag = 12)
```

ARIMA settles on a (2,1,0)(0,0,1) model with an AIC of 864, suggesting a better fit than the ETS model. This model uses Lagged CPI Growth as an additional predictor. There is some autocorrelation but it is not significant.

### Neural Network

```{r nn, warning=FALSE, message=FALSE}
# Create model
interest_rate_nn = interest_rate_train %>% 
  model(NNETAR(interest ~ CPI_lag))

report(interest_rate_nn)

interest_rate_nn %>% 
  gg_tsresiduals(lag_max = 12) +
  labs(title = "Neural Network Model Residuals")

# Run a Ljung_box test, value is not significant
augment(interest_rate_nn) %>% 
  features(.innov, ljung_box, lag = 12)
```

There are some issues with autocorrelation, which is significant at the 0.05 level, but I continue with the model. Transformations to the data created more significant autocorrelation.

## Prediction

I will move onto prediction with the three models created along with an ensemble model.

```{r predict, warning=FALSE, message=FALSE}
# Build final
interest_rate_model = interest_rate_train %>% 
  model(
    `ETS` = ETS(interest),
    `ARIMA` = ARIMA(interest ~ CPI_lag),
    `Neural Network` = NNETAR(interest ~ CPI_lag),
    `Ensemble` = combination_ensemble(ets = ETS(interest),
                                      arima = ARIMA(interest ~ CPI_lag),
                                      nn = NNETAR(interest ~ CPI_lag))
  )

# Forecast
interest_rate_forecast = interest_rate_model %>% 
  forecast(interest_rate_test)

# Plot
interest_rate_forecast %>% 
  autoplot(level = NULL, lwd = 1) +
  labs(x = "Date", y = "Interest Rate (%)",
       title = "Model Forecasts",
       subtitle = "Test Set is Black Line") + 
  guides(color = guide_legend(title = "Model")) + 
  geom_line(data = interest_rate_test, 
            aes(x = date, y = interest), 
            lwd = 1.25)

# Calculate accuracy
accuracy_table = interest_rate_forecast %>% 
  accuracy(interest_rate_test) %>% 
  select(-c(.type, MASE, RMSSE)) %>% 
  rename("Model" = ".model")

kable(accuracy_table,
      format = "markdown",
      caption = "Model Accuracy")

```

As seen above, none of the models particularly forecast well compared to previous models I have done in this course. The ARIMA model performs the best, but the MAPE is still 13.1%. The neural network performs the worst. Future work would include attempting other predictors for the data set along with other models (such as a TSLM). It is also worth noting that interest rates since the Great Recession post 2007 has been historically low as the economy has rebounded.